---
title: "Untargeted Metabolomics"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
})

# resolve conflicts
conflicted::conflict_prefer("filter", "dplyr")

# set global chunk options
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE, 
  fig.align = "center", 
  out.width = "49%"
) 

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  pos_chrom <- tar_read(chromatograms_pos)
  neg_chrom <- tar_read(chromatograms_neg)
  pos_int <- tar_read(intensities_pos)
  neg_int <- tar_read(intensities_neg)
})
```

\newpage

# Overview

We observed that proliferating primary cells exposed to hypoxia do not increase glucose uptake and lactate efflux despite up-regulation of glucose transporters and glycolytic genes. When these cells are treated with the prolyl hydroxylase inhibitor molidustat in normoxia, the expected increases in glycolytic flux are observed. Interestingly, when molidustat-treated cells are cultured in hypoxia, hypoxia blocks molidustat-mediated increases in glycolysis. In an effort to identify the mechanism mediating this effect, we performed metabolomics on lung fibroblasts treated for three days with 0.5% oxygen or molidustat (10 μM) with 21% and DMSO (0.1%) controls. 

# Data

## Preprocessing

The raw total ion intensity chromatogram were reviewed for an overview of signal acquisition quality:

```{r}
pos_chrom
neg_chrom
```

```{r}
pos_int
neg_int
```
